{\bf {\Large Parameter sensitivity and identifiability for a biogeochemical model of hypoxia in the northern {G}ulf of {M}exico}}

### 2016-12-06 ident_manu.Rnw
four pages done in discussion

Common goals in the application of biogeochemical models of ecosystem processes are to \begin{inparaenum}[1\upshape)]
\item accurately describe the system with historical data, and 
\item provide a means of forecasting ecosystem condition with hypothesized management or environmental scenarios. 
\end{inparaenum}
Although these objectives are the focus of most model applications, the structural components of process-based models should secondarily provide inference into which ecosystem processes and functions are driving observed or hypothesized changes.  This latter objective represents a more general scientific outcome of biogeochemical models that extends beyond the applied benefits of describing and predicting change in a particular system.  Modelers often hope to identify universal principles that govern dynamics across systems and the constraint of model parameters to observations provides a means of supporting or refuting these principles.  Extension of these principles to test the effects of structural changes and observation uncertainty on model predictions provideds further evidence to validate model components. This study provided a simple approach to use the effects of parameter perturbations on model state variables to characterize identifiable parameter subsets that vary by parameter selection criteria.  By doing so, we demonstrated that small parameter subsets relative to all sensitive parameters were within the identifiabilty thresholds described in the literature.  The identifiable parameter subsets varied considerably between state variables and the method for parameter selection.  We further demonstrated that changes in the model structure and variation in the initial conditions had an effect on sensitivity which has direct implications for identifiability. In general, these results provide justification for the use of explicit parameter selection heuristics that practitioners should adopt for model calibration.

Although the results were specific to individual variables, some generalities can be inferred from the sensitivity analsyes. State variables were most sensitive to parameters in the phytoplankton and zooplankton categories, particularly the maximum growth rates (\Sexpr{par_txt('umax_1')} for phytoplankton, \Sexpr{par_txt('Zumax_1')} for zooplankton), mortality coefficient for phytoplankton (\Sexpr{par_txt('mA_1')}), and the zooplankton half saturation coefficent for grazing (\Sexpr{par_txt('ZKa_1')}). An increase in growth rate of primary producers has the potential to increase oxygen concentration through photosynthetic processes, although increased production of organic matter is balanced with respiration and bacterial decomposition that reduce \ac{do} concentration in the water column.  Similarly, increases in zooplankton abundance with increased growth rates causes a reduction in phytoplankton biomass through grazing, which is expected to further deplete pool of organic matter.  Most variables were also sensitive to variation in the half-saturation grazing coefficient which is linked to the concentration of nutrients that support half the maximum grazing rate. Although the tradeoff between abundance, grazing, and decomposition is complex, the sensitivity of model state variables to parameters that directly control the abundance of primary producers is in agreement with empirical observations of factors that influence hypoxia dynamics on the \ac{lcs}. The sensitivity of the model output to variation in other parameters that relate to physical and chemical properties of the system was of secondary importance to biological relationships.  State variables were sensitive to changes in light and temperature parameters, although to a lesser extent than phytoplankton and zooplankton parameters.  As such, the differing sensitivities of state variables to parameters in each of the categories was not unexpected given general ecological relationships that are well understood and described by the model.

The overwhelming conclusion from the identifiablity analyses is that only limited subsets of parameters are identifiable within the constraints of local sensitivity analyses. Although we have not attempted actual model calibration (see recommendations below), these results support previous studies that have suggested similarly small subsets of parameters can be identified using traditional calibration schemes \citep[e.g.,][]{Wheater86,Ye97,Omlin01}.  In addition to \ac{cgem}, these conclusions have relevance for many biogeochemical models that include numerous parameters and structural equations to characterize processes in the model domain.  A general conclusion is that perhaps a trend towards less complex models could be beneficial given that only a small subset of parameters is identifiable and that ecosystem processes may in fact be sufficiently characterized with few parameters \citep{Ye97}.  Conversely, others have argued that model complexity is not in itself a disadvantage when parsimony is not the sole arbitrator of model structure \citep{Reichert97}. Over-parameterization can be useful if processes have importance that were not evaluated during model identification.  Our results demonstrated that approximately 75\% of the evaluated parameters had an effect on the eight state variables, whereas \ac{cgem} includes a total of 36 variables and multiple plankton groups, not all of which have immediate concern for understanding hypoxia. The redundancies identified with the sensitivity analyses are only problematic if the primary interest is, for example, \ac{do} dynamics.  Moreover, the proposed parameter selection heuristics provide flexibility for selecting different parameters based on parameter categories with the assumption that the chosen parameters depends on the research or management question.

% Commonalities in idenfitiability that emerged were 1) potentially driven by a single parameter, 2) nto related to sensitivity, 3) not always related to sheer number of aprmaters

%The \ac{cgem} model also has the ability to model multiple phytoplankton and zooplankton groups, whereas the FishTank model evaluated herein only evaluated one group each for simplicity.  The use of simple sensitivity analyses is insufficient to

% id of single-most redundant parameters - inclusion/exclusion or fig. 7

%' We showed only small subsets are identifiable, similar conclusions have been described by citations in \citep{Wagener01}, p. 14 for models that follow traditional calibration schemes (e.g., objective function minimization).
%' 
%' Emphasize that parameters that have the greatest effect on collinearity are not those that have the highest sensitivity (contrast the identifiability by category vs identifiability by top parameters)  , also note that groups of parameters together can have large effects on collinearity, maybe some kind of bootstrap analysis could be done looking at doubletons, etc. The example in teh results highlights how redundant variables can be identified as a necessary part of the model calibration process.
%' 
%' Why did identifiability decrease for top five parameters in each category after removing redundant phyto/zoop groups?
%' 
%' \citep{Denman03} describes over-parameterization for multiple phyto groups, use this to emphasize collinearity issues with the phyto groups despite differing sensitivity values between groups
%' 
%' Identifiability by category - varies with number of parameters in the category but some were more redundant than others (phytoplankton).  
%' 
%' Questions specific to GOM - what initial conditions are important? How many phytoplankton groups do we need (e.g., related to structural uncertainty)?
%' 
%' How does the assimilation of additional parameters (e.g., other state variables) during calibration influence the conclusions?  \citet{Wagener01} describes this as a potential approach to improving model performance by improving the availability of information for model calibration (p. 14).  
%' 
%' How does uncertainty translate to what a model should provide (generality v precision)?  The first step - find out what can be optimized but then do not overfit....
%' 
%' What about structural uncertainty - does sensitivity of a model to variation in a parameter imply parameter uncertainty and/or structural uncertainty?
%' 
%' A final point about optimization with identifiable parameter sets - optimization to fit the data still does not ensure a correct model.  Failing in one way can be over-compensated by another feature, e.g., the parameter set that is optimized (see \cite{Flynn05}, p. 1207, third paragraph), also \citep{Durand02,Arhonditsis08}, also note that over-parameterized models are not necessarily bad, see \citep{Omlin01}, models must be validated for the uses which they were intended where a comparison of observed and predicted is the simplest approach and other methods should be used depending on analysis needs \citep{Rykiel96}
%' 
%' \cite{Omlin01} state that the sensitivity, identifiability, estimation process is iterative (p. 113), need to rinse and repeat for proper calibration. 
%' 
%' How to improve identifiability - get more/better observed data, include obs from other state variables in RSS minimization (eqn q in \cite{Omlin01})
%' 
%' Alternative methods for uncertainty analysis - bayesian, MCMC, nonlinear calibration-constrained optimization \citep{Gallagher07}, \citep{Arhonditsis08}
%' 
%' Our heuristic products are partially analogous to the Rainfall-Runoff Modelling Toolbox (RRMT) presented by \citep{Wagener01b}2001 (cited on page 15, in \citealt{Wagener01}), although ours is simple in comparison

>>>>

\subsection{Implications of sensitivity and identifiability analyses}

Common goals in the application of biogeochemical models of ecosystem processes are to \begin{inparaenum}[1\upshape)]
\item accurately describe the system my matching predictions with observed data \citep{Reckhow90}, and 
\item provide a means of forecasting ecosystem condition with hypothesized management or environmental scenarios \citep{Clark01}. 
\end{inparaenum}
Although these objectives are the focus of most model applications, the structural components of process-based models should secondarily provide inference into which ecosystem processes and functions are driving observed or hypothesized changes.  This latter objective represents a more general scientific outcome of biogeochemical models that extends beyond the applied benefits of describing and predicting change in a particular system.  Modelers often hope to identify universal principles that govern dynamics across systems and the constraint of model parameters to observations provides a means of supporting or refuting hypotheses \citep{Kirchner06}.  Extension of these principles to test the effects of structural changes and observation uncertainty on model predictions provides further evidence to validate model components. This study provided a simple approach to use the effects of parameter perturbations on model state variables to characterize identifiable parameter subsets that vary by parameter selection criteria.  By doing so, we demonstrated that small parameter subsets relative to all sensitive parameters were within the identifiabilty thresholds described in the literature.  The identifiable parameter subsets varied considerably between state variables and the method for parameter selection.  We further demonstrated that changes in the model structure and variation in the initial conditions had an effect on sensitivity which has direct implications for identifiability. In general, these results provide justification for the use of explicit parameter selection heuristics that practitioners should adopt for model calibration.

Although the results were specific to individual variables, some generalities can be inferred from the sensitivity analyses. State variables were most sensitive to parameters in the phytoplankton and zooplankton categories, particularly the maximum growth rates (\Sexpr{par_txt('umax_1')} for phytoplankton, \Sexpr{par_txt('Zumax_1')} for zooplankton), mortality coefficient for phytoplankton (\Sexpr{par_txt('mA_1')}), and the zooplankton half saturation coefficent for grazing (\Sexpr{par_txt('ZKa_1')}). An increase in growth rate of primary producers has the potential to increase oxygen concentration through photosynthetic processes, although increased production of organic matter is balanced with respiration and bacterial decomposition that reduce \ac{do} concentration in the water column.  Similarly, increases in zooplankton abundance with increased growth rates causes a reduction in phytoplankton biomass through grazing, which is expected to further deplete pools of organic matter.  Most variables were also sensitive to variation in the half-saturation grazing coefficient which moderates the concentration of nutrients that support half the maximum grazing rate. Although the tradeoff between abundance, grazing, and decomposition is complex, the sensitivity of model state variables to parameters that directly control the abundance of primary producers is in agreement with empirical observations of factors that influence hypoxia dynamics on the \ac{lcs} \citep{Fahnenstiel95,Roelke00,Eldridge10}. The sensitivity of the model output to variation in other parameters that relate to physical and chemical properties of the system was of secondary importance to biological relationships.  State variables were sensitive to changes in light and temperature parameters, although to a lesser extent than phytoplankton and zooplankton parameters.  As such, the differing sensitivities of state variables to parameters in each of the categories was not unexpected given general ecological relationships that are well understood and described by the model.

The overwhelming conclusion from the identifiablity analyses is that only limited subsets of parameters are identifiable within the constraints of local sensitivity analyses. Although we have not attempted actual model calibration (see recommendations below), these results support previous studies that have suggested similarly small subsets of parameters can be identified using traditional calibration schemes \citep[e.g.,][]{Wheater86,Ye97,Omlin01}.  In addition to \ac{cgem}, these conclusions have relevance for many biogeochemical models that include numerous parameters and structural equations to characterize processes in the model domain.  A general conclusion is that perhaps a trend towards less complex models could be beneficial given that only a small subset of parameters is identifiable and that ecosystem processes may in fact be sufficiently characterized with few parameters \citep{Ye97}.  Conversely, others have argued that model complexity is not in itself a disadvantage when parsimony is not the sole arbitrator of model structure \citep{Reichert97}. Over-parameterization can be useful if processes have importance that were not evaluated during model identification.  Single objective functions that maximize model precision with identifiable parameters may also provide an incomplete characterization of model worth, which has prompted the development of probability-based models of hypoxia that explicitly include uncertainty in model components \citep[e.g.][]{Obenour15}.  Our results demonstrated that approximately 75\% of the evaluated parameters had an effect on the eight state variables, whereas \ac{cgem} includes a total of 36 variables and multiple plankton groups, not all of which have immediate concern for understanding hypoxia. The redundancies identified with the sensitivity analyses are only problematic if the primary interest is, for example, \ac{do} dynamics.  Moreover, the proposed parameter selection heuristics provide flexibility for selecting different parameters based on parameter categories with the assumption that the chosen parameters depends on the research or management question.

Results from the identifiability analyses provided additional insight into the interactions of parameters in large biogeochemical models.  First, identifiability of parameter subsets was not related to the sensitivity of individual variables. As noted above, an identifiable parameter is one that has a unique effect on model predictions that cannot be compensated for or undone by changing values of other parameters. The magnitude of the effect has no bearing on identifiability, which further complicates the selection of parameters for calibration.  Although identifiability is the primary limiting factor in choosing a given set, the relative sensitivities are more important for the decision to include or exclude indvidual parameters.  Our analysis addressed this challenge by presenting multiple selection criteria for identifiable parameter sets that prioritized the most sensitive parameters during the selection process.  Similarly, identifability was not always related to the number of parameters in a set. Although the overwhelming trend was decreasing identifiability with more parameters, the unique effects of including an individual parameter with an existing set often reduced the $\gamma$ estimate. For example, \cref{fig:exclex} shows that including \Sexpr{par_txt('KGcdom_1')}, \Sexpr{par_txt('KO2_1')}, or \Sexpr{par_txt('nitmax_1')} in parameters sets more often reduced $\gamma$ relative to sets that excluded the parameters.  Conversely, \cref{fig:heurist_stts} showed that the inclusion of a specific parame ters often caused a disporportionate increase in $\gamma$ (e.g., \Sexpr{par_txt('ediblevector(Z1)_1')}). These examples demonstrate the complex interactions of parameter changes on variable response, highlighting the need to consider the combined and individual effects of parameters on identifiability. The selection criteria proposed in our above analyses can facilitate parameter selection and also provide diagnostic tools to identify parameters with disproportionate effects on $\gamma$.

\subsection{Recommendations and conclusions}

An evaluation of parameter uncertainty and identifiability of relevant parameter sets is a preliminary and simplistic approach to improving model predictions.  In general, uncertainty analyses that lead to improved models are ultimately expected to increase our understanding of properties that define ecosystems using relationships described by the model.  The extension of simple parameter sensitivity analyses to the generalization of ecosystem properties requires additional analysis and potential model refinement, at the core of which is the balance between generality and precision. We emphasize that the utility of a specific model depends on the question and objective for application to a specific system.  For the above analysis, the FishTank model as part of the larger \ac{cgem} application was evaluated in the context of hypoxia effects on ecosystem condition and function.  At the core of the simple model is a set of biogeochemical equations that characterize the system in relation to planktonic groups, nutrient requirements, and water-column metabolism.  Our results have shown that relatively small subsets of parameters are identifiable given the complexity of the model, and as a result, we have provided a general approach to select parameter subsets depending on the ecological context (i.e., selection by parameter category, selection for specific state variables).  Thus, the results described above have relevance for further model refinement with the specific goal of better understanding ecological dynamics that moderate hypoxia on the northern \ac{gom}.  However, the general principles of sensitivity and parameter identifiability have broad applicability beyond this context and we argue that such methods should be more universally applied as an initial approach to quantify numerical constraints of biogeochemical models.

Specific approaches can be used to improve and build on the results presented herein, in addition to the more general considerations noted above. A logical next step is an evaluation of model precision following calibration with relevant parameter subsets.  For simplicity, our analysis did not calibrate model parameters and an explicit assumption was that parameter subsets with $\gamma$ below 10 or 15 were identifiable.  To our knowledge, this threshold has not been rigorously evaluated and a logical expectation is that this threshold is likely to vary between parameter subsets and the chosen method of model calibration.  Variation in parameter estimates given differences in the calibration method and identifiability thresholds has meaningful implications for intrerpreting model output.  Further, the extent to which the results for the simple \ac{zerod} extrapolate to the larger three-dimensional model should be evaluated.  Although the simple model

%The \ac{cgem} model also has the ability to model multiple phytoplankton and zooplankton groups, whereas the FishTank model evaluated herein only evaluated one group each for simplicity.  The use of simple sensitivity analyses is insufficient to
%' 
%' Emphasize that parameters that have the greatest effect on collinearity are not those that have the highest sensitivity (contrast the identifiability by category vs identifiability by top parameters)  , also note that groups of parameters together can have large effects on collinearity, maybe some kind of bootstrap analysis could be done looking at doubletons, etc. The example in teh results highlights how redundant variables can be identified as a necessary part of the model calibration process.
%' 
%' Questions specific to GOM - what initial conditions are important? How many phytoplankton groups do we need (e.g., related to structural uncertainty)?
%' 
%' How does the assimilation of additional parameters (e.g., other state variables) during calibration influence the conclusions?  \citet{Wagener01} describes this as a potential approach to improving model performance by improving the availability of information for model calibration (p. 14).  
%' 
%' How does uncertainty translate to what a model should provide (generality v precision)?  The first step - find out what can be optimized but then do not overfit....
%' 
%' What about structural uncertainty - does sensitivity of a model to variation in a parameter imply parameter uncertainty and/or structural uncertainty?
%' 
%' A final point about optimization with identifiable parameter sets - optimization to fit the data still does not ensure a correct model.  Failing in one way can be over-compensated by another feature, e.g., the parameter set that is optimized (see \cite{Flynn05}, p. 1207, third paragraph), also \citep{Durand02,Arhonditsis08}, also note that over-parameterized models are not necessarily bad, see \citep{Omlin01}, models must be validated for the uses which they were intended where a comparison of observed and predicted is the simplest approach and other methods should be used depending on analysis needs \citep{Rykiel96}
%' 
%' \cite{Omlin01} state that the sensitivity, identifiability, estimation process is iterative (p. 113), need to rinse and repeat for proper calibration. 
%' 
%' How to improve identifiability - get more/better observed data, include obs from other state variables in RSS minimization (eqn q in \cite{Omlin01})
%' 
%' Alternative methods for uncertainty analysis - bayesian, MCMC, nonlinear calibration-constrained optimization \citep{Gallagher07}, \citep{Arhonditsis08}
%' 
%' Our heuristic products are partially analogous to the Rainfall-Runoff Modelling Toolbox (RRMT) presented by \citep{Wagener01b}2001 (cited on page 15, in \citealt{Wagener01}), although ours is simple in comparison
